Implementing Capon Adaptive Beamforming on the GPU for Real Time Cardiac Ultrasound Imaging

### 2012-08-01 HRUIonGPU/jrnl_article.tex
small update

The solution to this optimization problem is given by the following expression:
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}},
\end{align}
where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed to broadside, and $\mat{\hat{R}}$ is the sample covariance matrix calculated as $\vec{x}[n]\vec{x}[n]^H$. 
%\begin{align}
%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
%\end{align} 
Note that this calculation only use a single sample. Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.

@@

The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
&= 1/N_L \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).

@@

\section{Parallel Capon}\label{sec:meth}
As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity. However, if we analyse the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$. As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that one block of threads should process at least one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes. That is, for each data vector $\vec{x}^\Delta$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively. As $L$ increases, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

The straight forward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ if $N_L$, the number of subarrays, is small. This is actually the same as for inverting a $L \times L$ matrix. However, little attention has been devoted to this step in the literature until now. Note that for a typical array configuration we typically do more spacial then temporal smoothing. Hence, $N_K$ is usually small compared to $N_L$. For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$, since $N_L = M - L + 1$. From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$. This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, if we want to avoid block-to-block synchronization.

The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time. The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+N_L-1}x_{j+N_L-1}^*$. The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-K-1]$ and adding $\mat{\hat{R}}[n+K]$. Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (figure?). However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$. Another aspect is that sliding will break independence in the sliding dimension. With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays. We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block. Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

\subsection{Solving Multiple Small Linear Systems}
In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries usually lack a comprehensive collection of batched solvers. The reason for this is two-fold, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases.

For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$. For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest MKL libraries.

@@

\section{Real Time Performance for $L > 16$}\label{sec:meth2}

\subsection{Beamspace Processing}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS} = \mat{B}\vec{x}^{\Delta}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space. We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$. For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy. Nilsen et. al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.

@@

\section{Results}\label{sec:res}

% stolpediagram, med de ulike metodenes kjøretid oppehverandre.

>>>>

The solution to this optimization problem is given by the following expression:
\begin{align}\label{eq:w}
\vec{w}[n] = \frac{\mat{\hat{R}}[n]^{-1}\vec{a}}{\vec{a}^H\mat{\hat{R}}[n]^{-1}\vec{a}} = \frac{\vec{b}}{\vec{a}^H\vec{b}},
\end{align}
where $\vec{a} = [1_0, 1_1, ..., 1_{M-1}]^T = \vec{1}_M$ when $\vec{x}$ is pre-delayed to broadside, and $\mat{\hat{R}}$ is the sample covariance matrix calculated as $\vec{x}[n]\vec{x}[n]^H$. 
%\begin{align}
%\mat{\hat{R}}[n] = \vec{x}[n]\vec{x}[n]^H.
%\end{align} 
Note that this calculation only use a single sample. Typically the sample covariance matrix has to be averaged over several samples in order to ensure numerical stability and statistical robustness.

@@

The beamformer output is in the end calculated by combining all $N_L$ subarrays weighted with the same set of adaptive weights.
\begin{align}
z[n] &= \frac{1}{N_L}\vec{w}[n]^H \sum_{l=0}^{N_L-1} \vec{x}_l^{\Delta}[n] \label{eq:z_mv}\\
&= 1/N_L \, \text{CONV}_m(\vec{w}[n], \vec{1}_{N_L})^H\vec{x}^{\Delta}[n] \label{eq:z_mv2},
\end{align}
%where (this matrix is wrong)
%\begin{align}
%\mat{C} = \left(
%\begin{matrix}
%\vec{1}_1 & \vec{1}_2 & \cdots & \vec{1}_{N_L} & \vec{0}_1 & \cdots & \vec{0}_{N_L-1} & \vec{0}_{N_L}\\
%\vec{0}_{N_L-1} & \vec{0}_{N_L-2} & \cdots & \vec{0}_{0} & \vec{1}_{N_L-1} & \cdots & \vec{1}_{2} & \vec{1}_{1}\\
%\end{matrix}
%\right)^T
%\end{align}
Note how we can either select to sum all subarrays first in (\ref{eq:z_mv}), or accumulate the right weighting per element in (\ref{eq:z_mv2}).

@@

\section{Parallel Capon}\label{sec:meth}
As mentioned, widespread use of Capon adaptive beamforming has been limited by its computational complexity. However, if we analyse the beamformer in (\ref{eq:z_mv}) in combination with (\ref{eq:w}) and (\ref{eq:R}), we see that each calculation of a set of Capon weights is independent across the output $z[n]$. As shown in Figure \ref{fig:gpulayout}, a first level of granularity is therefore to divide the image into a grid, where the output in each cell can be calculated independently of all others. Further, based on memory dependencies and to hold memory consumption per thread at a minimum, one should recognise that a group of threads should process one output in the final image. Continuing with the calculation of one weight vector, the algorithm can be broken down into three main steps that to some extent can be further divided into a small set of parallel processes. That is, for each data vector $\vec{x}^\Delta$ in the image:
\begin{align*}
%\begin{array}{l}
\text{1) } &\text{Calculate the sample covariance matrix } \mat{\hat{R}} \text{ as in } \text{(\ref{eq:diag})}\\
\text{2) } &\text{Solve the linear system of equations } \vec{b} =\mat{\hat{R}}^{-1}\vec{1}_L\\
\text{3) } &\text{Calculate } \vec{w} = \vec{b}/\vec{1}_{L}^H\vec{b} \text{ and the amplitude as in (\ref{eq:z_mv})}
%\end{array}
\end{align*}
%\begin{enumerate}
%\item Calculate the sample covariance matrix $\mat{\hat{R}}$ as in (\ref{eq:R})
%\end{enumerate}
%\begin{enumerate}
%\item[2)] Solve the linear system of equations $\mat{\hat{R}}\vec{b} =\vec{ 1}$
%\end{enumerate}
%\begin{enumerate}
%\item[3)] Calculate $\vec{w} = \vec{b}/\vec{1}^H\vec{b}$ and the amplitude $\vec{w}^H\vec{x}$
%\end{enumerate}
The computational complexity of these three steps is $O(L^3)$, $O(L^3)$, and $O(L)$ respectively. As $L$ increases, the first two steps quickly becomes unmanageable, and real time performance can only be achieved by exploring novel solutions to reduce complexity.

@@

The straight forward sum in (\ref{eq:R}) has complexity $O(L^2N_LN_K) = O(L^3)$ since $N_L$, the number of subarrays, is a function of $L$. This is actually the same as for inverting a $L \times L$ matrix. However little attention has been devoted to this step in the literature before. Note that for a typical array configuration we typically do more spacial then temporal smoothing. Hence, $N_K$ is usually small compared to $N_L$. For short subarrays the construction of $\mat{\hat{R}}$ is dominated by the number of elements $M$ ($N_L = M - L + 1$), but since $L$ is selected in most cases to be between $M/2$ and $M/4$ a large $M$ implies a large $L$.

From (\ref{eq:R}) it is obvious that calculation of one element in $\mat{\hat{R}}$ is independent of all the others. A natural level of granularity would then be to assign one thread to each element of $\mat{\hat{R}}$. This will minimize global reads and writes needed per thread, but unfortunately the maximum number of threads per block puts an upper limit on the subarray length, $L_{max}=32$, if we want to avoid block-to-block synchronization.

The quadratic complexity with $L$ and linear complexity with $N_K$ can be reduced by realizing that the calculation in (\ref{eq:R}) overlap both across subarrays and in time. The $(i,j)$ element of $\mat{\hat{R}}[n]$ can be calculated from element $(i-1, j-1), \text{for } i,j > 0$ by subtracting $x_{i-1}x_{j-1}^*$ and adding $x_{i+N_L-1}x_{j+N_L-1}^*$. The same is true in time where $\mat{\hat{R}}[n]$ can be calculated from $\mat{\hat{R}}[n-1]$ by subtracting $\mat{\hat{R}}[n-K-1]$ and adding $\mat{\hat{R}}[n+K]$. Both sliding across subarrays and in time can be realized simultaneously if we hold a $N_K$-average in time of the outer products $\vec{x}^{\Delta}[n]\vec{x}^{\Delta}[n]^H$ in memory and calculate $\mat{\hat{R}}$ averaged over subarrays from that buffer (ref JOE). However, since $N_K$ typically is small, and we have a limited amount of memory per thread on the GPU, we have chosen to focus on sliding across subarrays only. In that case, the complexity of calculating $\mat{\hat{R}}$ is reduced to $O(L^2)$. Important to notice is that sliding will break independence in the sliding dimension. With our choice of calculating $\mat{\hat{R}}$, calculations are still independent across time, but not across subarrays. We therefore have selected to use one thread per column in $\mat{\hat{R}}$, and for small $L$'s several covariance matrices are calculated per thread block (Figure \ref{fig:gpulayout}). Maximum supported $L$ is with this approach increased to 512, however for small $L$ high $M$ configurations, memory consumption per thread will be to high to achieve full occupancy of the GPU.

@@

\subsection{Solving Multiple Small Linear Systems}
In the literature, when it comes to solving systems of linear equations on the GPU the focus has mostly been on large systems, and GPU libraries therefore often lack a comprehensive collection of batched solvers. The reason for this is two-fold, first the GPU needs to solve thousand of small systems in order to beat the low memory latency of the CPU, second there is a range of system dimensions that does not fit well with the GPU architecture, where only a small amount of near-core memory and registers are available. Because of this, it has been proven hard to optimize a manageable set of solvers that provides speedups compared with the CPU in every cases. It is therefore not given that solving on the GPU is faster than solving on the CPU in general. However, there is two arguments for doing solving on the GPU. First the CPU, which already runs a lot of algorithms, is offloaded. Second the transfer of several thousand covariance matrices from CPU-side to GPU-side is spared.

For the results presented in this paper, we have used an unreleased GPU implementation of Gauss Jordan (GJ) elimination (by NVIDIA) to solve $\mat{\hat{R}}\vec{b} = \vec{1}_L$, where $\vec{b} = \mat{\hat{R}}^{-1}\vec{1}_L$. For smaller systems, $L < 10$, a direct solver using one thread per system is preferable. For this purpose we have implemented Cramer's rule for $2 \times 2$, $3 \times 3$ and $4 \times 4$ matrices. All solvers have been benchmarked against the latest Intel MKL libraries.

@@

\section{Beamspace Processing}
Nilsen et al. \cite{Nilsen2009} have proposed to apply beamspace processing to further reduce the required computations. With beamspace we mean transforming the channel data from element space to a fan of beams covering the sector illuminated by transmit beams. The transformation is written as
\begin{align}
\vec{x}_{BS} = \mat{B}\vec{x}^{\Delta}.
\end{align} 
The steering vectors in $\mat{B}$, the so-called Butler matrix, determines each beams direction in space. We can therefore easily remove beams where no interference is present by removing rows in $\mat{B}$. For a focused system like a cardiac ultrasound scanner, where the received signal is concentrated in a narrow band around broadside, there is a small percentage of the beams that contains almost all the energy. Nilsen et. al concluded that as little as three beams around broadside where adequate to produce results comparable to applying Capon beamforming in element space.

@@

\section{Results}\label{sec:res}
Describe the computer system (CPU, GPU...). Describe the data that are processed (field II, Channel data from GE Vingmed).
% stolpediagram, med de ulike metodenes kjøretid oppehverandre.
